---
title: "Study-Level Automation & Deep Features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Study-Level Automation & Deep Features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tlframe)
```

This guide covers automation and advanced features. For core pipeline verbs,
see `vignette("tlframe")`.

The central question here is: *"I can make one table. How do I make 50
efficiently?"*


## The Three-Tier Defaults System

tlframe resolves settings from four tiers, lowest to highest priority:

```
+----------------------------------------------------------------------+
| 4. Per-table verbs         fr_page(font_size = 10)          <- wins  |
| 3. Session theme           fr_theme(font_size = 9)                   |
| 2. Project config          _tlframe.yml                              |
| 1. Package defaults        inst/defaults/_tlframe.yml       <- lowest|
+----------------------------------------------------------------------+
```

Every `fr_table()` call merges all four tiers. You configure the broadest
settings at the lowest tier and override only what changes.

**Mental model:**

| Tier | Scope | Who sets it | When it changes |
|------|-------|-------------|-----------------|
| Package defaults | All users | tlframe maintainer | Never (ships with package) |
| `_tlframe.yml` | One study/project | Lead programmer | Once per study |
| `fr_theme()` | One R session | Programmer | Per program if needed |
| Per-table verbs | One table | Programmer | Per table |

> **SAS parallel:** Tier 2 is your `autoexec.sas` + `PROC TEMPLATE`. Tier 3
> is `OPTIONS` at the top of your program. Tier 4 is inline `STYLE()=`
> overrides.


## YAML Configuration: Company House Style

Place a `_tlframe.yml` at your project root. Every table in the project
inherits these settings without any R code.

### Full file reference

```yaml
# _tlframe.yml — place at project root
page:
  paper: letter                       # letter | a4 | legal
  orientation: landscape              # landscape | portrait
  margins: [1.0, 0.75, 1.0, 0.75]    # top, right, bottom, left (inches)
  font_family: "Courier New"          # Any system font
  font_size: 9                        # Points
  continuation: "(continued)"         # Appended to titles on page 2+
  col_gap: 4                          # Inter-column gap in points

columns:
  split: false                        # true = auto-split wide tables

header:
  align: ~                            # null = inherit from column
  valign: bottom                      # top | middle | bottom
  bold: false
  span_gap: true                      # Gap between adjacent spans
  n_format: "{label}\n(N={n})"         # N-count label template

pagehead:
  left: "Protocol TFRM-2024-001"
  center: ~
  right: "CONFIDENTIAL"
  font_size: ~                        # null = inherit page font_size
  bold: false

pagefoot:
  left: "{program}"
  center: ~
  right: "Page {thepage} of {total_pages}"
  font_size: ~
  bold: false

rules:
  hlines: header                      # header | open | booktabs | box | void
  vlines: void                        # box | all | inner | void

footnotes:
  separator: false                    # true = horizontal line above footnotes
  font_size: ~
  placement: every                    # every | last (last = PDF only)

spacing:
  titles_after: 1                     # Blank lines after titles
  footnotes_before: 1                 # Blank lines before footnotes
  pagehead_after: 0
  pagefoot_before: 0
  page_by_after: 1

titles:
  align: center                       # left | center | right
  bold: false
  font_size: ~

tokens:
  study: "TFRM-2024-001"
  compound: "Zomerane"
  cutoff: "15MAR2025"
```

### Loading config

```{r config-demo}
# Auto-discover: searches up from getwd() for _tlframe.yml
# Falls back to package defaults if not found
fr_config()

# Explicit path
# fr_config("path/to/_tlframe.yml")

# Inspect current config
cfg <- fr_config_get()
cfg$page$font_size

# Reset to package defaults
fr_config_reset()
```

### Config-driven tokens

Custom tokens defined in YAML are available in `fr_pagehead()` and
`fr_pagefoot()`:

```yaml
tokens:
  study: "TFRM-2024-001"
  cutoff: "15MAR2025"
```

```{r tokens-config, eval = FALSE}
fr_config("_tlframe.yml")

# Now {study} and {cutoff} resolve automatically
tbl_demog |>
  fr_table() |>
  fr_pagehead(left = "Study {study}", right = "Cutoff: {cutoff}") |>
  fr_render("output.rtf")
```

No R code needed to define the token values — they come from the YAML file.


## Theme System: Study-Level Overrides

`fr_theme()` sets session-level defaults that override config and are
overridden by per-table verbs. It follows the `ggplot2::theme_update()`
pattern: call it once at the top of your program.

### Setting a theme

```{r theme-set}
fr_theme(
  font_size   = 9,
  font_family = "Courier New",
  orientation = "landscape",
  hlines      = "header",
  pagehead    = list(left = "TFRM-2024-001", right = "CONFIDENTIAL"),
  pagefoot    = list(left = "{program}",
                     right = "Page {thepage} of {total_pages}"),
  tokens      = list(study = "TFRM-2024-001"),
  spacing     = list(titles_after = 1, footnotes_before = 1)
)

# Verify
theme <- fr_theme_get()
theme$font_size
theme$pagehead$right
```

### Every table inherits the theme

```{r theme-inherit}
# Both tables inherit the theme automatically — no repeated configuration
spec1 <- tbl_demog |>
  fr_table() |>
  fr_titles("Table 14.1.1", "Demographics")

spec2 <- tbl_disp |>
  fr_table() |>
  fr_titles("Table 14.1.4", "Subject Disposition")

fr_get_page(spec1)$font_size
fr_get_page(spec2)$font_size
```

### Per-table override

```{r theme-override}
# This table needs portrait orientation — override just that
spec3 <- tbl_ae_soc |>
  fr_table() |>
  fr_page(orientation = "portrait") |>
  fr_titles("Table 14.3.1", "AE by SOC/PT")

fr_get_page(spec3)$orientation
fr_get_page(spec3)$font_size  # Still 9 from theme
```

### Reset

```{r theme-reset}
fr_theme_reset()
# Back to package defaults
is.null(fr_theme_get()$font_size)
```

### Theme vs config

| Feature | `_tlframe.yml` (config) | `fr_theme()` |
|---------|------------------------|--------------|
| Storage | File on disk | R session memory |
| Scope | All programs in project | Current R session |
| Set by | `fr_config()` | `fr_theme()` |
| Priority | Lower (tier 2) | Higher (tier 3) |
| Survives restart | Yes (file persists) | No (session-only) |
| Best for | Company/study-wide standards | Program-specific overrides |


## Recipes: Reusable Pipeline Fragments

Recipes capture `fr_*()` verb chains as portable objects. Unlike themes,
recipes can include column definitions, styles, and spans — anything you
can pipe.

### Creating recipes

```{r recipe-create}
# Company-wide page layout
company_layout <- fr_recipe(
  fr_page(orientation = "landscape", font_size = 9),
  fr_pagehead(left = "TFRM-2024-001", right = "CONFIDENTIAL"),
  fr_pagefoot(left = "{program}",
              right = "Page {thepage} of {total_pages}"),
  fr_hlines("header")
)
company_layout

# AE-specific formatting
ae_style <- fr_recipe(
  fr_header(bold = TRUE, align = "center"),
  fr_page(continuation = "(continued)")
)
ae_style
```

### Applying recipes

```{r recipe-apply}
spec <- tbl_demog |>
  fr_table() |>
  fr_apply(company_layout) |>
  fr_titles("Table 14.1.1", "Demographics")

fr_get_page(spec)$font_size
```

### Composing recipes

`c()` merges recipes. Later recipes override earlier ones for verbs that
replace; verbs that append (`fr_spans`, `fr_styles`) accumulate.

```{r recipe-compose}
# Combine company layout + AE-specific overrides
ae_combined <- c(company_layout, ae_style)

spec <- tbl_ae_soc |>
  fr_table() |>
  fr_apply(ae_combined) |>
  fr_titles("Table 14.3.1", "AE by SOC/PT")

fr_get_page(spec)$continuation
```

### The Three-Layer Recipe Pattern

For a study with 50+ tables, structure recipes in three layers:

```{r three-layer, eval = FALSE}
# Layer 1: Company standard (shared across all studies)
company <- fr_recipe(
  fr_page(orientation = "landscape", font_family = "Courier New",
          font_size = 9),
  fr_pagehead(right = "CONFIDENTIAL"),
  fr_pagefoot(left = "{program}",
              right = "Page {thepage} of {total_pages}"),
  fr_hlines("header")
)

# Layer 2: Study-specific (shared across all tables in one study)
study <- fr_recipe(
  fr_pagehead(left = "Protocol TFRM-2024-001"),
  fr_page(tokens = list(study = "TFRM-2024-001",
                         cutoff = "15MAR2025"))
)

# Layer 3: Table-type recipes
demog_recipe <- fr_recipe(
  fr_header(bold = TRUE, align = "center",
            n = c(placebo = 45, zom_50mg = 45,
                  zom_100mg = 45, total = 135),
            format = "{label}\n(N={n})"),
  fr_footnotes("Percentages based on N per treatment arm.",
               .separator = FALSE)
)

ae_recipe <- fr_recipe(
  fr_header(bold = TRUE, align = "center",
            n = c(placebo = 45, zom_50mg = 45,
                  zom_100mg = 45, total = 135),
            format = "{label}\n(N={n})"),
  fr_page(continuation = "(continued)"),
  fr_footnotes("MedDRA version 26.0.",
               "Subjects counted once per SOC and PT.",
               .separator = FALSE)
)

# Usage: compose layers + add per-table details
tbl_demog |>
  fr_table() |>
  fr_apply(c(company, study, demog_recipe)) |>
  fr_titles("Table 14.1.1", "Demographics") |>
  fr_cols(group = fr_col(visible = FALSE)) |>
  fr_render("output/Table_14_1_1.rtf")

tbl_ae_soc |>
  fr_table() |>
  fr_apply(c(company, study, ae_recipe)) |>
  fr_titles("Table 14.3.1", "AE by SOC/PT") |>
  fr_cols(soc = fr_col(visible = FALSE),
          row_type = fr_col(visible = FALSE)) |>
  fr_rows(group_by = "soc", indent_by = "pt") |>
  fr_render("output/Table_14_3_1.rtf")
```

### Saving and loading recipes

Recipes are serializable — they store R call expressions, not closures:

```{r recipe-serialise, eval = FALSE}
saveRDS(company, "company_style.rds")
company <- readRDS("company_style.rds")
```

This means recipes can be shared across team members via a shared drive or
internal R package.


## Study Automation Blueprint

For a typical CSR with 50--100 TLFs, combine a metadata table with recipes
and a batch loop.

### Step 1: Create a metadata CSV

```{r metadata-csv}
metadata <- data.frame(
  table_id  = c("14.1.1", "14.1.4", "14.3.1", "14.2.1"),
  title1    = c("Demographics", "Subject Disposition",
                "AE by SOC/PT", "Time to Study Withdrawal"),
  title2    = c("ITT Population", "All Randomized",
                "Safety Population", "ITT Population"),
  data_name = c("tbl_demog", "tbl_disp", "tbl_ae_soc", "tbl_tte"),
  recipe    = c("demog", "demog", "ae", "tte"),
  stringsAsFactors = FALSE
)
metadata
```

### Step 2: Define recipe library

```{r recipe-lib, eval = FALSE}
recipes <- list(
  demog = fr_recipe(
    fr_header(bold = TRUE, align = "center",
              n = c(placebo = 45, zom_50mg = 45,
                    zom_100mg = 45, total = 135)),
    fr_footnotes("Percentages based on N per arm.", .separator = FALSE)
  ),
  ae = fr_recipe(
    fr_header(bold = TRUE, align = "center",
              n = c(placebo = 45, zom_50mg = 45,
                    zom_100mg = 45, total = 135)),
    fr_page(continuation = "(continued)"),
    fr_rows(group_by = "soc", indent_by = "pt"),
    fr_cols(soc = fr_col(visible = FALSE),
            row_type = fr_col(visible = FALSE)),
    fr_footnotes("MedDRA v26.0.", .separator = FALSE)
  ),
  tte = fr_recipe(
    fr_header(bold = TRUE, align = "center",
              n = c(zom_50mg = 45, zom_100mg = 45, placebo = 45)),
    fr_rows(group_by = "section", blank_after = "section"),
    fr_cols(section = fr_col(visible = FALSE)),
    fr_footnotes("[a] Kaplan-Meier estimate.", .separator = FALSE)
  )
)
```

### Step 3: Batch rendering loop

```{r batch-loop, eval = FALSE}
company_study <- c(company, study)  # From the three-layer pattern above

for (i in seq_len(nrow(metadata))) {
  row <- metadata[i, ]

  # Look up the pre-summarized dataset by name
  data <- get(row$data_name)

  # Look up the table-type recipe
  recipe <- recipes[[row$recipe]]

  # Build and render
  data |>
    fr_table() |>
    fr_apply(c(company_study, recipe)) |>
    fr_titles(
      paste("Table", row$table_id),
      row$title1,
      row$title2
    ) |>
    fr_render(sprintf("output/Table_%s.rtf",
                      gsub("\\.", "_", row$table_id)))
}
```

The entire 4-table study output is generated with a 6-line loop. Scale to
50 tables by adding rows to the metadata CSV.

> **SAS parallel:** This replaces the SAS macro library + `%include` pattern.
> The metadata CSV is your macro call dataset, recipes are your macros, and
> the loop is your `%DO` block.


## Column Splitting for Wide Tables

When a table has more columns than fit on a page, `fr_cols(.split = TRUE)`
automatically splits into panels:

```{r col-split-demo}
spec <- tbl_vs[tbl_vs$timepoint == "Week 24" &
                tbl_vs$param == "Systolic BP (mmHg)", ] |>
  fr_table() |>
  fr_titles("Table 14.3.6", "Systolic BP --- Column Split Demo") |>
  fr_cols(
    param     = fr_col(visible = FALSE),
    timepoint = fr_col(visible = FALSE),
    statistic = fr_col("Statistic", width = 1.2, stub = TRUE),
    .split = TRUE, .width = "fit",
    placebo_base      = fr_col("Placebo\nBaseline", width = 1.0),
    placebo_value     = fr_col("Placebo\nValue", width = 1.0),
    placebo_chg       = fr_col("Placebo\nCFB", width = 1.0),
    zom_50mg_base     = fr_col("Zom 50mg\nBaseline", width = 1.0),
    zom_50mg_value    = fr_col("Zom 50mg\nValue", width = 1.0),
    zom_50mg_chg      = fr_col("Zom 50mg\nCFB", width = 1.0),
    zom_100mg_base    = fr_col("Zom 100mg\nBaseline", width = 1.0),
    zom_100mg_value   = fr_col("Zom 100mg\nValue", width = 1.0),
    zom_100mg_chg     = fr_col("Zom 100mg\nCFB", width = 1.0)
  ) |>
  fr_hlines("header")
```

**Key concepts:**

- **Stub columns** (`stub = TRUE`) repeat in every panel. Use for row labels.
- **`.split = TRUE`** enables automatic panel splitting when total width
  exceeds the printable area.
- **`.width = "fit"`** scales each panel to fill the page width.
- Panel widths are calculated proportionally — wider columns get more space.


## Complex Styling Patterns

### Conditional styling with fr_style_if

`fr_style_if()` applies styles based on data values at render time. Unlike
`fr_rows_matches()` which selects rows, `fr_style_if()` evaluates a
condition per cell:

```{r style-if-demo}
spec <- tbl_demog |>
  fr_table() |>
  fr_cols(group = fr_col(visible = FALSE)) |>
  fr_styles(
    # Bold all group header rows
    fr_row_style(
      rows = fr_rows_matches("group", pattern = "^[A-Z]"),
      bold = TRUE
    ),
    # Tint the total column
    fr_col_style(cols = "total", bg = "#F0F4F8")
  )
```

### Style precedence

When multiple styles target the same cell, the last one wins. Order your
`fr_styles()` calls from broadest to most specific:

```{r style-precedence}
spec <- tbl_demog |>
  fr_table() |>
  fr_cols(group = fr_col(visible = FALSE)) |>
  fr_styles(
    # 1. Base: tint total column (broadest)
    fr_col_style(cols = "total", bg = "#F0F4F8"),
    # 2. Bold group headers
    fr_row_style(
      rows = fr_rows_matches("group", pattern = "^[A-Z]"),
      bold = TRUE
    ),
    # 3. Specific cell override (most specific — wins)
    fr_style(rows = 1, cols = "total", bg = "#003366", fg = "#FFFFFF")
  )
```

### Zebra striping

```{r zebra-demo}
spec <- tbl_disp |>
  fr_table() |>
  fr_styles(
    fr_row_style(rows = seq(2, nrow(tbl_disp), 2), bg = "#F5F5F5")
  ) |>
  fr_hlines("header")
```

### Header region styling

```{r header-region}
spec <- tbl_demog |>
  fr_table() |>
  fr_styles(
    fr_style(region = "header", bold = TRUE,
             bg = "#003366", fg = "#FFFFFF", align = "center")
  ) |>
  fr_hlines("header")
```


## RTF vs PDF: What Differs

The same spec produces both formats, but some features behave differently:

| Feature | RTF | PDF |
|---------|-----|-----|
| Rendering engine | Direct RTF 1.9.1 control words | tabularray + XeLaTeX |
| Footnote placement `"last"` | Not supported (repeats every page) | Works (lastfoot) |
| Page numbering | `\chpgn` / `\field` codes | `\thepage` / `\pageref` |
| Continuation text | Plain text on page 2+ | LaTeX `\fancyhead` |
| Decimal alignment | `\ql\li<offset>` (left + indent) | `\hspace{pt}` + `~` |
| Fonts | Referenced by name (viewer uses local) | Embedded via fontspec |
| Multi-line cells | `\line` (RTF line break) | `\newline` (LaTeX) |
| Group pagination | R-side pagination (with `group_by`) | `longtblr` + `\trkeep` |

**Font handling on Linux:**

RTF files reference fonts by name — the viewing application uses whatever
fonts are installed locally. PDF files embed fonts via XeLaTeX's fontspec.
On bare Linux systems without Microsoft fonts, tlframe automatically falls
back to bundled Liberation fonts:

| System font | Liberation fallback |
|-------------|-------------------|
| Courier New | Liberation Mono |
| Arial | Liberation Sans |
| Times New Roman | Liberation Serif |

### PDF requirements

PDF rendering requires XeLaTeX. Install via tinytex:

```{r pdf-setup, eval = FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

Or use your system's TeX distribution. tlframe compiles twice for correct
page references.


## QC Patterns

### Pre-render validation

```{r qc-validate}
spec <- tbl_demog |>
  fr_table() |>
  fr_titles("Table 14.1.1", "Demographics") |>
  fr_cols(group = fr_col(visible = FALSE))

result <- fr_validate(spec)
result$valid
```

`fr_validate()` checks:

- Column names match the data frame
- `group_by`, `page_by`, `indent_by` reference valid columns
- Style row/column indices are in range
- Widths are positive and reasonable
- Span columns exist and are contiguous

Use `strict = TRUE` in CI pipelines to fail on warnings.

### Accessor-based QC checks

The `fr_get_*()` functions enable programmatic quality checks:

```{r qc-accessors}
spec <- tbl_demog |>
  fr_table() |>
  fr_titles("Table 14.1.1", "Demographics") |>
  fr_footnotes("Source: ADSL.", .separator = FALSE) |>
  fr_cols(group = fr_col(visible = FALSE))

# Check title count
stopifnot(length(fr_get_titles(spec)) == 2)

# Check footnote content
fn <- fr_get_footnotes(spec)[[1]]$content
stopifnot(grepl("ADSL", fn))

# Check page defaults
pg <- fr_get_page(spec)
stopifnot(pg$orientation == "landscape")
stopifnot(pg$font_size == 9)

# Check visible columns
cols <- fr_get_columns(spec)
visible <- names(cols)[vapply(cols, function(x) {
  vis <- x$visible
  if (is.null(vis)) TRUE else vis
}, logical(1))]
stopifnot("group" %in% setdiff(names(cols), visible))
```

### Style debugging

`fr_style_explain()` shows the complete resolution chain for a style
object, which is useful when debugging overlapping styles:

```{r qc-style-debug}
spec <- tbl_demog |>
  fr_table() |>
  fr_styles(
    fr_row_style(rows = 1, bold = TRUE),
    fr_col_style(cols = "total", bg = "#EBF5FB")
  )
fr_style_explain(spec, row = 1, col = "total")
```

### Double-programming workflow

For regulatory submissions, the QC programmer can build a parallel spec
and compare:

```{r qc-double, eval = FALSE}
# Programmer 1: creates the table
spec1 <- tbl_demog |>
  fr_table() |>
  fr_titles("Table 14.1.1", "Demographics") |>
  fr_render("output/t_14_1_1_prod.rtf")

# Programmer 2 (QC): independently creates the same table
spec2 <- tbl_demog |>
  fr_table() |>
  fr_titles("Table 14.1.1", "Demographics") |>
  fr_render("output/t_14_1_1_qc.rtf")

# Compare: diff the RTF files or use tools::Rdiff()
# Identical RTF = verified
```

The immutable pipeline makes double-programming straightforward — same
data + same verbs = identical output.


## What's Next

For architecture details, the render pipeline, custom backends, and the
sentinel system: `vignette("extending-tlframe")`.
