---
title: "Architecture & Custom Backends"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Architecture & Custom Backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tlframe)
```

This guide covers tlframe's internal architecture. It is intended for
developers who want to understand the render pipeline, write custom
backends, or contribute to the package.


## Architecture Overview

tlframe is organised into three layers across 24 source modules:

```
+----------------------------------------------------------------------+
| API Layer (10 files)                                                 |
| fr_table -> fr_cols -> fr_header -> fr_titles -> ... -> fr_render    |
| Each verb: validate -> modify fr_spec -> return fr_spec              |
+----------------------------------------------------------------------+
| Infrastructure Layer (9 files)                                       |
| classes.R     S3 constructors (fr_spec, fr_col, fr_rule, ...)        |
| columns.R     Width estimation, AFM metrics, auto-distribution       |
| config.R      YAML discovery, loading, merging                       |
| constants.R   fr_env: fonts, paper sizes, presets, colors            |
| helpers.R     Error conditions, text normalisation                   |
| markup.R      Inline rich text: fr_super() -> sentinel tokens        |
| tokens.R      {thepage}, {program}, {datetime} resolution            |
| units.R       Twips, half-points, row height calculation             |
| validate.R    Type-safe validation helpers                           |
+----------------------------------------------------------------------+
| Render Layer (5 files)                                               |
| render.R          Dispatch: finalize -> prepare_pages -> backend     |
| render-common.R   Shared: cell grid, borders, escaping, colors       |
| render-rtf.R      RTF 1.9.1 backend                                  |
| render-latex.R    tabularray/XeLaTeX backend                         |
| render-pdf.R      PDF wrapper (generates .tex, compiles twice)       |
+----------------------------------------------------------------------+
```

### The fr_spec object

Every pipeline verb reads and writes to a single S3 object, `fr_spec`.
Its key fields:

| Field | Type | Set by |
|-------|------|--------|
| `$data` | data.frame | `fr_table()` |
| `$meta$titles` | list of title entries | `fr_titles()` |
| `$meta$footnotes` | list of footnote entries | `fr_footnotes()` |
| `$columns` | named list of `fr_col` | `fr_cols()` |
| `$columns_meta` | list (split, width_mode) | `fr_cols()` |
| `$header` | list (n, format, bold, ...) | `fr_header()` |
| `$body` | list (page_by, group_by, ...) | `fr_rows()` |
| `$rules` | list of `fr_rule` | `fr_hlines()`, `fr_vlines()` |
| `$cell_styles` | list of `fr_cell_style` | `fr_styles()` |
| `$page` | list (orientation, margins, ...) | `fr_page()` |
| `$pagehead` | list (left, center, right) | `fr_pagehead()` |
| `$pagefoot` | list (left, center, right) | `fr_pagefoot()` |
| `$spacing` | list (titles_after, ...) | `fr_spacing()` |

### The fr_env environment

`fr_env` (in `constants.R`) is a private package environment holding all
lookup data and mutable state:

```{r fr-env}
# Font families and their AFM font names
names(fr_backends())

# Paper sizes are stored internally
pg <- fr_get_page(tbl_demog |> fr_table())
pg$paper
```

Key contents: `$fonts` (family → AFM mapping), `$paper_sizes` (name →
dimensions), `$hline_presets` (name → rule set), `$named_colors` (148 CSS
colors), `$unicode_map` (symbol → codepoint), `$liberation_fallback`
(system font → Liberation font).


## The Render Pipeline

When `fr_render(spec, path)` is called, it runs a three-stage pipeline:

```
fr_spec
  |
  v
finalize_spec()            <- Resolve all deferred computations
  |  - Build default columns for unmentioned data columns
  |  - Distribute widths (auto/fit/equal/percent)
  |  - Resolve N-count header labels
  |  - Insert blank_after rows
  |  - Apply indent_by offsets
  |  - Compute decimal geometry for align="decimal" columns
  |
  v
prepare_pages()            <- Split data into page groups
  |  - Split by page_by column values (preserving data order)
  |  - Each group: {label, data, n_counts}
  |
  v
calculate_col_panels()     <- Handle wide tables
  |  - If .split=TRUE and total width > page: split into panels
  |  - Stub columns repeat in each panel
  |  - Scale panel widths proportionally
  |
  v
render_<format>()          <- Backend-specific rendering
     RTF:   render_rtf(spec, page_groups, col_panels, path)
     LaTeX: render_latex(spec, page_groups, col_panels, path)
     PDF:   render_pdf() -> render_latex() + xelatex compilation
```

### finalize_spec: The resolution stage

This is where deferred computations are resolved. Key operations:

**Width distribution:** If `.width = "auto"`, column widths are estimated
from content using AFM font metrics. If total width exceeds the printable
area, columns are scaled down proportionally:

```{r finalize-width}
spec <- tbl_demog |>
  fr_table() |>
  fr_cols(.width = "auto", group = fr_col(visible = FALSE))

# Widths are resolved at render time, not when fr_cols() is called
# This allows verb-order independence
```

**N-count labels:** `fr_header(n = ..., format = "{name}\n(N={n})")` stores
the N-counts and format template. `finalize_spec()` resolves them into final
column labels so that `fr_cols()` and `fr_header()` can be called in any
order.

**Blank rows:** `blank_after = "group"` inserts empty rows after each group
boundary. `group_by` auto-implies `blank_after` on the same column.

**Decimal geometry:** For `align = "decimal"` columns, the 15-type stat
display engine analyses each cell value, detects its type (count, float,
n/%, CI, p-value, range, etc.), and computes space-padded alignment offsets.

### prepare_pages: The splitting stage

When `page_by = "param"` is set, data is split by `unique(data$param)`
(preserving original data order, not alphabetical). Each page group is
rendered as a separate section.

### calculate_col_panels: The width stage

When `.split = TRUE` and the total column width exceeds the printable area,
columns are distributed across panels. Stub columns repeat in every panel.
Each panel's widths are scaled proportionally to fill the page.


## Writing a Custom Backend

tlframe's backend interface is formalised. Any backend must implement a
single function with this signature:

```r
render_<format>(spec, page_groups, col_panels, path)
```

**Arguments:**

| Argument | Type | Description |
|----------|------|-------------|
| `spec` | Finalised `fr_spec` | All deferred computations resolved |
| `page_groups` | List of lists | Each: `$label`, `$data`, `$n_counts` |
| `col_panels` | List of char vectors | Column name groups per panel |
| `path` | Character scalar | Output file path |

**Return:** The output file path (character scalar).

### Registering a backend

```{r register-backend, eval = FALSE}
# Register a custom backend for HTML output
render_html <- function(spec, page_groups, col_panels, path) {
  # ... generate HTML ...
  writeLines(html, path)
  path
}

fr_register_backend("html", render_html, extensions = "html")

# Now this works:
spec |> fr_render("output.html")
```

### Shared utilities in render-common.R

The render layer provides shared utilities that any backend can use:

**`build_cell_grid()`** — Converts the data frame into a matrix of rendered
cell content, applying inline markup resolution and decimal alignment.

**`build_header_cell_grid()`** — Same for the header row, with N-count
labels already resolved.

**`resolve_borders()`** — Converts the `$rules` list into a per-cell border
specification (top, bottom, left, right for each cell).

**`collect_colors()`** — Scans all styles and backgrounds to build a color
table (RTF needs a `\colortbl`; LaTeX needs `\definecolor` preamble).

**`rtf_escape_and_resolve()` / `latex_escape_and_resolve()`** — Escape
special characters and resolve markup sentinels (`\x01SUPER:a\x02` →
`{\super a}` or `\textsuperscript{a}`).


## The Markup Sentinel System

Inline markup (`fr_super()`, `fr_bold()`, etc.) works through a sentinel
token system that survives string concatenation:

```
User writes:       "P{fr_super('a')}-value < 0.05"
                          ↓
glue evaluation:   "P\x01SUPER:a\x02-value < 0.05"
                          ↓
RTF backend:       "P{\super a}-value < 0.05"
LaTeX backend:     "P\textsuperscript{a}-value < 0.05"
```

The `\x01...\x02` sentinel tokens are non-printable bytes that:

1. Survive `paste()`, `sprintf()`, and other string operations
2. Are invisible in console output
3. Are resolved per-backend at the final rendering stage

This design means markup works in titles, footnotes, column labels,
and data cells — anywhere a string is accepted.

### Supported sentinel types

| Type | Sentinel | RTF | LaTeX |
|------|----------|-----|-------|
| Superscript | `\x01SUPER:x\x02` | `{\super x}` | `\textsuperscript{x}` |
| Subscript | `\x01SUB:x\x02` | `{\sub x}` | `\textsubscript{x}` |
| Bold | `\x01BOLD:x\x02` | `{\b x}` | `\textbf{x}` |
| Italic | `\x01ITALIC:x\x02` | `{\i x}` | `\textit{x}` |
| Underline | `\x01UNDERLINE:x\x02` | `{\ul x}` | `\underline{x}` |
| Newline | `\x01NEWLINE\x02` | `\line` | `\newline` |
| Unicode | `\x01UNICODE:code\x02` | `\u<code>?` | Direct UTF-8 |


## AFM Font Metrics

tlframe uses real Adobe Font Metrics for accurate column width estimation.
Pre-parsed metrics for 12 font variants are stored in `R/sysdata.rda`:

| Family | Regular | Bold | Italic | Bold Italic |
|--------|---------|------|--------|-------------|
| Helvetica | Helvetica | Helvetica-Bold | Helvetica-Oblique | Helvetica-BoldOblique |
| Times | Times-Roman | Times-Bold | Times-Italic | Times-BoldItalic |
| Courier | Courier | Courier-Bold | Courier-Oblique | Courier-BoldOblique |

**Width ratios illustrate why AFM matters:**

- Helvetica 'M' = 833 units vs 'i' = 222 units (3.75x ratio)
- Courier is fixed at 600 units for all characters
- A "Summary Statistics" column in Helvetica needs different width than in
  Courier, and tlframe accounts for this automatically

The `estimate_col_width()` function measures actual text widths (both header
labels and data content) using `measure_text_width_twips()`, which sums
per-character AFM widths for the correct font variant.


## CI/CD and Docker

### Batch rendering in CI

```{r ci-batch, eval = FALSE}
# render_all.R — run in CI pipeline
library(tlframe)
fr_config("_tlframe.yml")

metadata <- read.csv("metadata.csv", stringsAsFactors = FALSE)
for (i in seq_len(nrow(metadata))) {
  row <- metadata[i, ]
  data <- get(row$data_name)
  recipe <- readRDS(sprintf("recipes/%s.rds", row$recipe))

  data |>
    fr_table() |>
    fr_apply(recipe) |>
    fr_titles(paste("Table", row$table_id), row$title1) |>
    fr_render(sprintf("output/Table_%s.rtf",
                      gsub("\\.", "_", row$table_id)))
}
```

### Docker setup

For reproducible PDF rendering in containers:

```dockerfile
FROM rocker/r-ver:4.3.3

# Install system dependencies
RUN apt-get update && apt-get install -y \
    libxml2-dev libcurl4-openssl-dev libssl-dev

# Install R packages
RUN R -e "install.packages(c('cli','glue','rlang','tidyselect','vctrs','yaml','stringi'))"
RUN R -e "install.packages('remotes'); remotes::install_github('vthanik/tlframe')"

# For PDF output: install TinyTeX
RUN R -e "install.packages('tinytex'); tinytex::install_tinytex()"

# Liberation fonts are bundled with tlframe — no Microsoft fonts needed
# tlframe auto-detects and uses Liberation as fallback on Linux
```

No Microsoft fonts required in the container — tlframe's bundled Liberation
fonts provide pixel-compatible rendering for Courier New, Arial, and
Times New Roman.

### Validation in CI

```{r ci-validate, eval = FALSE}
# validate_all.R — run after rendering
library(tlframe)

metadata <- read.csv("metadata.csv", stringsAsFactors = FALSE)
errors <- character(0)

for (i in seq_len(nrow(metadata))) {
  row <- metadata[i, ]
  data <- get(row$data_name)
  recipe <- readRDS(sprintf("recipes/%s.rds", row$recipe))

  spec <- data |>
    fr_table() |>
    fr_apply(recipe) |>
    fr_titles(paste("Table", row$table_id), row$title1)

  result <- fr_validate(spec, strict = FALSE)
  if (!result$valid) {
    errors <- c(errors, sprintf("Table %s: %s",
                                 row$table_id,
                                 paste(result$messages, collapse = "; ")))
  }
}

if (length(errors) > 0) {
  stop("Validation failures:\n", paste(errors, collapse = "\n"))
}
```
