---
title: "Table Cookbook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Table Cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tlframe)
```

Complete, copy-paste-ready table programs using built-in datasets. Each
recipe follows ICH E3 table numbering.

## 14.1.1 Demographics

```{r demog}
demog_spec <- tbl_demog |>
  fr_table() |>
  fr_titles(
    "Table 14.1.1",
    "Demographics and Baseline Characteristics",
    "Intent-to-Treat Population"
  ) |>
  fr_cols(
    .width = "fit",
    characteristic = fr_col("", width = 2.5),
    placebo        = fr_col("Placebo", align = "decimal"),
    zom_50mg       = fr_col("Zomerane 50mg", align = "decimal"),
    zom_100mg      = fr_col("Zomerane 100mg", align = "decimal"),
    total          = fr_col("Total", align = "decimal"),
    group          = fr_col(visible = FALSE),
    .n = c(placebo = 45, zom_50mg = 45, zom_100mg = 45, total = 135),
    .n_format = "{label}\n(N={n})"
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_rows(group_by = "group", blank_after = "group") |>
  fr_hlines("header") |>
  fr_footnotes(
    "Percentages based on number of subjects per treatment group.",
    "MMSE = Mini-Mental State Examination.",
    .separator = FALSE
  )
fr_validate(demog_spec)$valid
```

```{r demog-render, eval = FALSE}
demog_spec |> fr_render("output/Table_14_1_1.rtf")
demog_spec |> fr_render("output/Table_14_1_1.pdf")
```

## 14.1.4 Subject Disposition

```{r disp}
disp_spec <- tbl_disp |>
  fr_table() |>
  fr_titles(
    "Table 14.1.4",
    list("Subject Disposition", bold = TRUE),
    "All Randomized Subjects"
  ) |>
  fr_cols(
    category  = fr_col("", width = 2.5),
    placebo   = fr_col("Placebo", align = "decimal"),
    zom_50mg  = fr_col("Zomerane 50mg", align = "decimal"),
    zom_100mg = fr_col("Zomerane 100mg", align = "decimal"),
    total     = fr_col("Total", align = "decimal"),
    .n = c(placebo = 45, zom_50mg = 45, zom_100mg = 45, total = 135),
    .n_format = "{label}\n(N={n})"
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_hlines("header") |>
  fr_footnotes(
    "Percentages based on number of subjects randomized per arm.",
    .separator = FALSE
  )
fr_validate(disp_spec)$valid
```

## 14.3.1 AE by SOC/PT

Multi-page table with continuation text, SOC/PT indentation, and
content-based row styling:

```{r ae-soc}
ae_soc_spec <- tbl_ae_soc |>
  fr_table() |>
  fr_titles(
    "Table 14.3.1",
    list("Treatment-Emergent Adverse Events by SOC and Preferred Term",
         bold = TRUE),
    "Safety Population"
  ) |>
  fr_pagehead(left = "TFRM-2024-001", right = "CONFIDENTIAL") |>
  fr_pagefoot(left = "{program}",
              right = "Page {thepage} of {total_pages}") |>
  fr_page(continuation = "(continued)") |>
  fr_cols(
    soc       = fr_col(visible = FALSE),
    pt        = fr_col("System Organ Class\n  Preferred Term", width = 3.5),
    row_type  = fr_col(visible = FALSE),
    placebo   = fr_col("Placebo", align = "decimal"),
    zom_50mg  = fr_col("Zomerane\n50mg", align = "decimal"),
    zom_100mg = fr_col("Zomerane\n100mg", align = "decimal"),
    total     = fr_col("Total", align = "decimal"),
    .n = c(placebo = 45, zom_50mg = 45, zom_100mg = 45, total = 135),
    .n_format = "{label}\n(N={n})"
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_rows(group_by = "soc", indent_by = "pt") |>
  fr_hlines("header") |>
  fr_styles(
    fr_row_style(
      rows = fr_rows_matches("row_type", value = "total"), bold = TRUE),
    fr_row_style(
      rows = fr_rows_matches("row_type", value = "soc"), bold = TRUE)
  ) |>
  fr_footnotes(
    "MedDRA version 26.0.",
    "Subjects counted once per SOC and Preferred Term.",
    "Sorted by descending total incidence.",
    .separator = FALSE
  )
fr_validate(ae_soc_spec)$valid
```

## 14.2.1 Time-to-Event

```{r tte}
tte_spec <- tbl_tte |>
  fr_table() |>
  fr_titles(
    "Table 14.2.1",
    list("Time to Study Withdrawal", bold = TRUE),
    "Intent-to-Treat Population"
  ) |>
  fr_cols(
    section   = fr_col(visible = FALSE),
    statistic = fr_col("", width = 3.5),
    zom_50mg  = fr_col("Zomerane\n50mg", align = "decimal"),
    zom_100mg = fr_col("Zomerane\n100mg", align = "decimal"),
    placebo   = fr_col("Placebo", align = "decimal"),
    .n = c(zom_50mg = 45, zom_100mg = 45, placebo = 45),
    .n_format = "{label}\n(N={n})"
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_rows(group_by = "section", blank_after = "section") |>
  fr_hlines("header") |>
  fr_styles(
    fr_row_style(
      rows = fr_rows_matches("statistic", pattern = "^[A-Z]"),
      bold = TRUE
    )
  ) |>
  fr_footnotes(
    "[a] Kaplan-Meier estimate with Greenwood 95% CI.",
    "[b] Two-sided log-rank test stratified by age group.",
    "[c] Cox proportional hazards model.",
    "NE = Not Estimable.",
    .separator = FALSE
  )
fr_validate(tte_spec)$valid
```

## 14.4.1 Concomitant Medications

```{r cm}
cm_spec <- tbl_cm |>
  fr_table() |>
  fr_titles(
    "Table 14.4.1",
    list("Concomitant Medications by Category and Agent", bold = TRUE),
    "Safety Population"
  ) |>
  fr_cols(
    category   = fr_col(visible = FALSE),
    medication = fr_col("Medication Category / Agent", width = 3.0),
    row_type   = fr_col(visible = FALSE),
    placebo    = fr_col("Placebo", align = "decimal"),
    zom_50mg   = fr_col("Zomerane\n50mg", align = "decimal"),
    zom_100mg  = fr_col("Zomerane\n100mg", align = "decimal"),
    total      = fr_col("Total", align = "decimal"),
    .n = c(placebo = 45, zom_50mg = 45, zom_100mg = 45, total = 135),
    .n_format = "{label}\n(N={n})"
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_rows(group_by = "category", indent_by = "medication") |>
  fr_hlines("header") |>
  fr_styles(
    fr_row_style(
      rows = fr_rows_matches("row_type", value = "total"), bold = TRUE),
    fr_row_style(
      rows = fr_rows_matches("row_type", value = "category"), bold = TRUE)
  ) |>
  fr_footnotes(
    "Subjects counted once per category and medication.",
    .separator = FALSE
  )
fr_validate(cm_spec)$valid
```

## 14.3.6 Vital Signs with page_by

Multi-parameter table with spanning headers and per-page N-counts via
a function:

```{r vs}
# Pre-compute per-parameter N-counts from ADVS
vs_n <- aggregate(
  USUBJID ~ PARAM + TRTA, data = advs[advs$AVISIT == "Baseline", ],
  FUN = function(x) length(unique(x))
)

vs_spec <- tbl_vs[tbl_vs$timepoint == "Week 24", ] |>
  fr_table() |>
  fr_titles(
    "Table 14.3.6",
    "Vital Signs --- Week 24 Summary",
    "Safety Population"
  ) |>
  fr_cols(
    param     = fr_col(visible = FALSE),
    timepoint = fr_col(visible = FALSE),
    statistic         = fr_col("Statistic", width = 1.2),
    placebo_base      = fr_col("Baseline"),
    placebo_value     = fr_col("Value"),
    placebo_chg       = fr_col("CFB"),
    zom_50mg_base     = fr_col("Baseline"),
    zom_50mg_value    = fr_col("Value"),
    zom_50mg_chg      = fr_col("CFB"),
    zom_100mg_base    = fr_col("Baseline"),
    zom_100mg_value   = fr_col("Value"),
    zom_100mg_chg     = fr_col("CFB"),
    .n = vs_n,
    .n_format = "{label}\n(N={n})"
  ) |>
  fr_rows(page_by = "param") |>
  fr_header(bold = TRUE, align = "center") |>
  fr_spans(
    "Placebo"        = c("placebo_base", "placebo_value", "placebo_chg"),
    "Zomerane 50mg"  = c("zom_50mg_base", "zom_50mg_value", "zom_50mg_chg"),
    "Zomerane 100mg" = c("zom_100mg_base", "zom_100mg_value", "zom_100mg_chg")
  ) |>
  fr_hlines("header") |>
  fr_footnotes("CFB = Change from Baseline.", .separator = FALSE)
fr_validate(vs_spec)$valid
```

## Wide table with column split

When columns exceed the page width, `.split = TRUE` creates panels:

```{r wide}
wide_spec <- tbl_vs[tbl_vs$timepoint == "Week 24" &
                     tbl_vs$param == "Systolic BP (mmHg)", ] |>
  fr_table() |>
  fr_titles("Table 14.3.6", "Systolic BP --- Column Split") |>
  fr_cols(
    param     = fr_col(visible = FALSE),
    timepoint = fr_col(visible = FALSE),
    statistic = fr_col("Statistic", width = 1.2, stub = TRUE),
    .split = TRUE, .width = "fit",
    placebo_base      = fr_col("Placebo\nBaseline", width = 1.0),
    placebo_value     = fr_col("Placebo\nValue", width = 1.0),
    placebo_chg       = fr_col("Placebo\nCFB", width = 1.0),
    zom_50mg_base     = fr_col("Zom 50mg\nBaseline", width = 1.0),
    zom_50mg_value    = fr_col("Zom 50mg\nValue", width = 1.0),
    zom_50mg_chg      = fr_col("Zom 50mg\nCFB", width = 1.0),
    zom_100mg_base    = fr_col("Zom 100mg\nBaseline", width = 1.0),
    zom_100mg_value   = fr_col("Zom 100mg\nValue", width = 1.0),
    zom_100mg_chg     = fr_col("Zom 100mg\nCFB", width = 1.0)
  ) |>
  fr_hlines("header")
fr_validate(wide_spec)$valid
```
