---
title: "Get Started with tlframe"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started with tlframe}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tlframe)
```

tlframe produces regulatory-grade clinical tables in RTF and PDF from a
single pipeline. You describe *what* the table looks like; tlframe handles
the rendering. No dplyr, no gt, no RMarkdown --- just data in, document out.

## Your first table

```{r first}
spec <- tbl_demog |>
  fr_table() |>
  fr_titles("Table 14.1.1", "Demographics and Baseline Characteristics") |>
  fr_hlines("header")
spec
```

That's a complete, renderable table. To produce a file:

```{r render-first, eval = FALSE}
spec |> fr_render("Table_14_1_1.rtf")
spec |> fr_render("Table_14_1_1.pdf")
```

Same spec, both formats. The file extension controls the output.

## Add structure

Real pharma tables need column labels, N-counts, footnotes, and page chrome.
Here's the same table, production-ready:

```{r production}
spec <- tbl_demog |>
  fr_table() |>
  fr_titles(
    "Table 14.1.1",
    "Demographics and Baseline Characteristics",
    "Intent-to-Treat Population"
  ) |>
  fr_cols(
    .width = "fit",
    characteristic = fr_col("", width = 2.5),
    placebo        = fr_col("Placebo", align = "decimal"),
    zom_50mg       = fr_col("Zomerane 50mg", align = "decimal"),
    zom_100mg      = fr_col("Zomerane 100mg", align = "decimal"),
    total          = fr_col("Total", align = "decimal"),
    group          = fr_col(visible = FALSE),
    .n = c(placebo = 45, zom_50mg = 45, zom_100mg = 45, total = 135),
    .n_format = "{label}\n(N={n})"
  ) |>
  fr_header(bold = TRUE, align = "center") |>
  fr_rows(group_by = "group", blank_after = "group") |>
  fr_hlines("header") |>
  fr_pagehead(left = "Protocol TFRM-2024-001", right = "CONFIDENTIAL") |>
  fr_pagefoot(left = "{program}", right = "Page {thepage} of {total_pages}") |>
  fr_footnotes(
    "Percentages based on number of subjects per treatment group.",
    .separator = FALSE
  )
fr_validate(spec)$valid
```

**12 lines of verbs** --- that's a submission-ready demographics table with
N-counts, decimal alignment, grouped rows, page headers/footers, and
footnotes. Every verb is optional; add only what you need.

## Table anatomy

Every pharma table has the same structural regions. Here's how they map to
tlframe verbs:

```
+------------------------------------------------------------------------+
| Protocol TFRM-2024-001                                    CONFIDENTIAL |<- fr_pagehead()
+------------------------------------------------------------------------+
|                                                                        |
|                           Table 14.1.1                                 |<- fr_titles()
|               Demographics and Baseline Characteristics                |
|                                                                        |
| ---------------------------------------------------------------------- |
|                  Placebo    Zomerane 50mg   Zomerane 100mg    Total     |<- fr_cols() +
|                   (N=45)      (N=45)          (N=45)        (N=135)    |   fr_cols(.n=)
| ====================================================================== |<- fr_hlines()
| Age (years)                                                            |
|   Mean (SD)    68.2 (7.1)   67.8 (6.9)     68.0 (7.3)    68.0 (7.1)  |<- data rows
|                                                                        |
| Sex, n (%)                                                             |<- fr_rows(blank_after=)
|   Male         26 (57.8)    24 (53.3)      25 (55.6)     75 (55.6)   |
| ---------------------------------------------------------------------- |
| Percentages based on number of subjects per treatment group.           |<- fr_footnotes()
+------------------------------------------------------------------------+
| t_demog.R                                                 Page 1 of 1  |<- fr_pagefoot()
+------------------------------------------------------------------------+
```

10 verbs, 10 regions. Learn the verbs, build any table.

## The pipeline

Every verb takes a spec, returns a spec. Verb order doesn't matter ---
tlframe resolves everything (widths, N-count labels, page breaks) at
render time.

```r
data |>
  fr_table()    |>   # create spec
  fr_titles()   |>   # add titles
  fr_cols()     |>   # configure columns, N-counts
  fr_header()   |>   # style the header row
  fr_rows()     |>   # row grouping, pagination
  fr_hlines()   |>   # horizontal rules
  fr_styles()   |>   # cell-level formatting
  fr_render("out.rtf")  # produce output
```

## What's next

| I want to... | Read |
|--------------|------|
| Configure columns, widths, decimal alignment | [Columns & Headers](columns-and-headers.html) |
| Add N-counts to column headers | [Columns & Headers](columns-and-headers.html) |
| Write titles, footnotes, superscripts | [Titles & Footnotes](titles-and-footnotes.html) |
| Group rows, paginate, indent SOC/PT | [Rows & Page Layout](rows-and-pages.html) |
| Add borders, spanning headers, cell colors | [Rules, Spans & Styles](styling.html) |
| Copy a complete table recipe | [Table Cookbook](table-cookbook.html) |
| Create listings or embed figures | [Listings & Figures](listings-and-figures.html) |
| Set up study-wide defaults, batch render | [Automation & Batch](automation.html) |
| Understand the internals | [Architecture](architecture.html) |
